---
title: "Prioritizing Vernal Pool Conservation In Central Massachusetts"
author: "Michael Jordan"
fontsize: 11pt
header-includes: 
 \usepackage{geometry}
 \geometry{top=2.5cm,left=2.5cm,bottom=2.5cm,right=2.5cm}
output: 
  pdf_document:
    latex_engine: xelatex
mainfont: Times New Roman
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

Massachusetts state law protects vernal pools from development. However, this protection is contingent on a pool being certified, which can only be requested at the discretion of private citizens. To help private citizens prioritize pool certification requests, this paper uses cluster analysis to identify hot spots of high conservation value pools in MassGIS potential pool data.

## Introduction
```{r import, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(terra)
library(sf)
library(units)
library(scales)
library(factoextra)
library(knitr)
library(cowplot)

setwd("C:/Users/Michael Jordan/Desktop/VernalPoolMiniProject")

# Import vectors MA towns, certified pools, potential pools, and wetlands.
ma.towns <- sf::st_read("./townssurvey_shp/TOWNSSURVEY_POLY.shp")
certified.pools <- sf::st_zm(
  sf::st_read("./GISDATA_CVP_PT/GISDATA_CVP_PTPoint.shp"))
potential.pools <- sf::st_read("./pvp/PVP_PT.shp")
wetlands <- sf::st_read("./wetlandsdep/WETLANDSDEP_POLY.shp")

# Create vectors for just the towns of interest in Central MA.
cm.towns <- ma.towns[
  ma.towns$TOWN %in% c(
    "BERLIN", "BOLTON", "HUDSON"), ]
cm.cp <- certified.pools[unlist(
  unique(
    sf::st_intersects(
      x = sf::st_geometry(cm.towns),
      y = sf::st_geometry(certified.pools)
    ))), ]
cm.pp <- potential.pools[unlist(
  unique(
    sf::st_intersects(
      x = sf::st_geometry(cm.towns),
      y = sf::st_geometry(potential.pools)
    ))), ]
cm.wetlands <- wetlands[unlist(
  unique(
    sf::st_intersects(
      x = sf::st_geometry(cm.towns),
      y = sf::st_geometry(wetlands)
    ))), ]

# Import land use tiles that intersect with towns. To do this, import tile
# indexes, convert to MA projection, get IDs of tiles overlapping with towns,
# create a list to hold tiles, import tiles by name, bind tiles together, and
# convert to MA projection.
tile.index <- sf::st_read(
  "./landcover_use_index_poly/LANDCOVER_USE_INDEX_POLY.shp")
tile.index <- sf::st_transform(tile.index, 26986)
needed.tiles <- tile.index[unlist(
  unique(
    sf::st_intersects(
      x = sf::st_geometry(cm.towns),
      y = sf::st_geometry(tile.index)
    ))), ]$TILENAME
tiles.l <- vector("list", length = length(needed.tiles))
for (i in 1:length(needed.tiles)) {
  shp <- sf::st_read(
    paste("./tiles/LCLU_",
      needed.tiles[i],
      "/LCLU_",
      needed.tiles[i],
      ".shp",
      sep = ""))
  tiles.l[[i]] <- shp[c("COVERCODE", "USEGENCODE")]
}
cm.tiles <- do.call(rbind, tiles.l)
cm.tiles <- sf::st_transform(cm.tiles, 26986)

# Filter tiles to include only polygons intersecting with towns.
cm.tiles <- cm.tiles[unlist(
  unique(
    sf::st_intersects(
      x = sf::st_geometry(cm.towns),
      y = sf::st_geometry(cm.tiles)
    ))), ]

```

Vernal pools are ephemeral bodies of water, often geographically separate from other wetlands, that typically fill in the spring and evaporate by summer (Burne & Griffin, 2005). Because vernal pools are not filled continuously, they do not support breeding fish populations, which makes them critical habitat for amphibian and invertebrate species that cannot successfully reproduce in the presence of fish predation (Burne & Griffin, 2005). In the state of Massachusetts, such obligate habitators of vernal pools include the wood frog (*Lithobates sylvaticus*), several species of mole salamanders (*Ambystoma* spp.), and various fairy shrimp species (*Eubranchipus* spp.) (MA DFG, 2009). Without vernal pools, these species would be under high risk of local extinction and others that use the pools on a facultative basis would face additional pressure.

Because of their significance as critical wildlife habitat, vernal pools are protected under the Massachusetts Wetlands Protection Act (WPA), which establishes that development is not permitted within 100 feet of a vernal pool's boundary (i.e., water line) (MA DFW, 2009). However, a pool must meet two criteria to qualify for WPA protection. First, it must exist within 100 feet of a wetland under WPA jurisdiction (MA DFW, 2009). Second, it must be certified by the Natural Heritage and Endangered Species Program (NHESP) as functioning biologically as a vernal pool (MA DFW, 2009). Fortunately, WPA wetland jurisdiction is fairly broad. But, the requirement that vernal pools be certified does significantly constrain their protection.

The reason for this is that the Massachusetts government does not actively pursue vernal pool certification, but instead relies on private citizens to do so (MA DFW, 2009). If a citizen would like to request certification of a potential vernal pool, they must visit the pool, collect biological data demonstrating the presence of certain species, and submit them to the NHESP. The lack of an active certification process has likely left tens of thousands of potential vernal pools unreviewed per MassGIS data. This number of pools presents a prioritization problem. Which pools should interested citizens review first?

In this analysis, I aim to help focus vernal pool conservation in Massachusetts by identifying potential vernal pools that should be prioritized for review because they are more likely to receive WPA protection, face higher threat from development, or can be reviewed more efficiently based on their geographic clustering.

## Methods

```{r first_plot, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Plot pools and wetlands versus town boundaries for visual inspection.
terra::plot(sf::st_geometry(cm.towns), col = "white")
terra::plot(
  sf::st_geometry(cm.wetlands),
  add = TRUE,
  col = "blue")
terra::plot(
  sf::st_geometry(cm.cp),
  add = TRUE,
  col = "green",
  pch = 17,
  cex = .5)
terra::plot(
  sf::st_geometry(cm.pp),
  add = TRUE,
  col = "red",
  pch = 17,
  cex = .5)
```
I completed this analysis using `r substring(R.Version()$version.string, 1, 15)` (see [Rmd](https://github.com/Skirnir3141/VernalPoolMiniProject/blob/master/MichaelJordanVernalPoolMiniProject.Rmd) for full code). To keep the analysis computationally tractable, I focused on three towns in Central Massachusetts: Berlin, Bolton, and Hudson. I downloaded vectors for [municipal boundaries](https://www.mass.gov/info-details/massgis-data-municipalities), [potential vernal pools](https://www.mass.gov/info-details/massgis-data-nhesp-potential-vernal-pools), [certified vernal pools](https://www.mass.gov/info-details/massgis-data-nhesp-certified-vernal-pools), [wetland boundaries](https://www.mass.gov/info-details/massgis-data-massdep-wetlands-2005), and [land cover/use](https://www.mass.gov/info-details/massgis-data-2016-land-coverland-use) from MassGIS on January 18, 2024. Pools were represented as geo points identifying the center of a pool and are dated to 1993-2000. Wetlands and land cover/use were represented as polygons and are dated to 2005 and 2016 respectively. Because the potential pool data set was not mutually exclusive from the certified pool data set, potential pools that are likely to represent certified pools were excluded based on proximity to certified pools (65ft threshold based on 15ft precision in potential pools and 50ft precision in certified pools) (Bourne, 2001). Vectors were plotted in the NAD83 / Massachusetts Mainland projection and visually inspected for errors.

```{r impervious_features, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Create an object including only polygons describing terrain that could
# indicate a threat to pools.  These are: COVERCODE 2 = Impervious, USEGENCODE 4
# = Industrial, USEGENCODE 10 = Mixed use, primarily residential, USEGENCODE 11
# = Residential - single family, USEGENCODE 12 = Residential - multi-family,
# USEGENCODE 30 = Mixed use, primarily commercial, USEGENCODE 55 = Right-of-way.
cm.tiles.f <- cm.tiles[
  cm.tiles$COVERCODE %in% c(2) |
    cm.tiles$USEGENCODE %in% c(4, 10, 11, 12, 30, 55), ]

# Check if pools overlap with an impervious feature. If they do, set their
# distance to the nearest impervious feature equal to zero. If they don't,
# calculate their distance to the nearest impervious feature in feet.
cm.pp$within.feature <- lengths(
  sf::st_intersects(
    sf::st_geometry(cm.pp),
    sf::st_geometry(cm.tiles.f)))
cm.pp.wf <- cm.pp[cm.pp$within.feature == 1, ]
cm.pp.wf$dtf <- 0
cm.pp.nwf <- cm.pp[cm.pp$within.feature == 0, ]
dist <- vector("list", length = nrow(cm.pp.nwf))
for (i in 1:nrow(cm.pp.nwf)) {
  dist[i] <- units::set_units(
    min(
      sf::st_distance(
        sf::st_geometry(cm.pp.nwf)[i],
        sf::st_geometry(cm.tiles.f))),
    'ft')
}
cm.pp.nwf$dtf <- dist
cm.pp <- rbind(cm.pp.nwf, cm.pp.wf)
```

To proxy for the threat each pool faces from human development, I measured the distance from each potential pool to the nearest land cover/use vector that could be expected to threaten the biological functioning of a pool (e.g., roads or industrial buildings). I defined this set of vectors based on NOAA C-CAP land cover/use codes (see [Rmd](https://github.com/Skirnir3141/VernalPoolMiniProject/blob/master/MichaelJordanVernalPoolMiniProject.Rmd) for code mapping). I refer to this as the set of "impervious features."

```{r biology, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Create an object that combines certified and potential pools
cm.pp.temp <- cm.pp["PVP_NUMBER"]
colnames(cm.pp.temp)[1] <- "pool_num"
cm.pp.temp$type <- "potential"
cm.cp.temp <- cm.cp["cvp_num"]
colnames(cm.cp.temp)[1] <- "pool_num"
cm.cp.temp$type <- "certified"
cm.p <- rbind(cm.pp.temp, cm.cp.temp)

# Run hierarchical cluster analysis. Note that units here are in meters.
cm.p.dist.matrix <- sf::st_distance(cm.p, cm.p)
cm.p.clust <- hclust(as.dist(cm.p.dist.matrix), method = "single")
# Add cluster cuts.
# The first is for the dispersal networks 150m is the average dispersal distance
# from the McDonough & Paton study. 38m is added to account for the distance
# from potential pool center to boundary edge based on 125ft being the diameter
# of pool easily discernible in the Bourne aerial photograph study (125 ft / 2
# ~= 38m).
cm.p$clust1 <- cutree(cm.p.clust, h = (150 + 38))
# The second is to disentangle potential vernal pools that are already in the
# certified vernal pool data set. Here the threshold is set to 20m ~= 15ft +
# 50ft i.e., the precision of each data set.
cm.p$clust2 <- cutree(cm.p.clust, h = 20)

# Get clusters containing both a certified and potential vernal pool according
# to both thresholds.
agg.cm.p.clust1 <- sf::st_drop_geometry(cm.p) %>%
  group_by(clust1) %>%
  summarise(types = toString(sort(unique(type))))
hv.clust1 <- agg.cm.p.clust1[
  agg.cm.p.clust1$types == "certified, potential", ]$clust1
agg.cm.p.clust2 <- sf::st_drop_geometry(cm.p) %>%
  group_by(clust2) %>%
  summarise(types = toString(sort(unique(type))))
hv.clust2 <- agg.cm.p.clust2[
  agg.cm.p.clust2$types == "certified, potential", ]$clust2


# Add columns for whether a potential pool is in a cluster with a certified
# pool versus both clusters thresholds.
cm.pp$bio.flag <- cm.pp$PVP_NUMBER %in% cm.p[
  cm.p$clust1 %in% hv.clust1, ]$pool_num
cm.pp$repeat.flag <- cm.pp$PVP_NUMBER %in% cm.p[
  cm.p$clust2 %in% hv.clust2, ]$pool_num

# Drop potential pools that are repeats based on proximity to a certified pool.
cm.pp <- cm.pp[cm.pp$repeat.flag == FALSE, ]
```

To proxy for the likelihood that a potential pool contains certifiable species, I used hierarchical cluster analysis with a single linkage fusion strategy to identify networks of pools that were no farther from each other than the expected dispersal distance of one such species, the spotted salamander (*Ambystoma maculatum*). I set this distance at 150m based on the average dispersal distance of spotted salamanders in a radio telemetry study in Connecticut (McDonough & Paton, 2010). The idea behind this approach is that pools connected to each other within this distance could sustain a metapopulation that travels between pools. 

```{r wetlands, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Check whether a potential pool overlaps with a wetland. If it does, set its
# distance to the nearest wetland equal to zero. If it doesn't, calculate its
# distance to the nearest wetland in feet.
cm.pp$within.wetland <- lengths(
  sf::st_intersects(cm.pp, cm.wetlands))
cm.pp.ww <- cm.pp[cm.pp$within.wetland == 1, ]
cm.pp.ww$dtw <- 0
cm.pp.nww <- cm.pp[cm.pp$within.wetland == 0, ]
dist2 <- vector("list", length = nrow(cm.pp.nww))
for (i in 1:nrow(cm.pp.nww)) {
  dist2[i] <- units::set_units(
    min(
      sf::st_distance(
        sf::st_geometry(cm.pp.nww)[i],
        sf::st_geometry(cm.wetlands))),
    'ft')
}
cm.pp.nww$dtw <- dist2
cm.pp <- rbind(cm.pp.nww, cm.pp.ww)

# Create an object that drops potential pools that are more than 115 feet from a
# wetland. 62.5ft are added to account for the distance from potential pool
# center to boundary edge based on 125ft being the diameter of pool easily discernible in the Bourne aerial photograph study (125 ft / 2).
cm.pp.f <- cm.pp[cm.pp$dtw <= (100 + 15 + 125 / 2), ]
```

Because the boundary of a pool must exist within 100 feet of a wetland for the pool to receive WPA protection, I evaluated each pool's proximity to a wetland. Potential pools that were within 115 feet of the nearest wetland were deemed to be candidates for certification. The rest were excluded. This cut off was based on the 100 foot WPA jurisdiction threshold, plus 15 feet of imprecision in potential pool vectors (Burne, 2001).

```{r cluster, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Run cluster analysis with a single linkage fusion strategy based on a geodesic
# distance of 1/4 of a mile. This distance is arbitrary, but the idea is that
# if volunteers have to drive too far between pools, the grouping doesn't
# represent a contiguous cluster that can be efficiently evaluated. Note that
# distance is left in meters, 402m ~= .25 miles.
cm.pp.f.dist.matrix <- sf::st_distance(cm.pp.f, cm.pp.f)
cm.pp.f.clust <- hclust(as.dist(cm.pp.f.dist.matrix), method = "single")
cm.pp.f$clust <- cutree(cm.pp.f.clust, h = 402)
```

To flag potential pools that are geographically close together and so could be reviewed efficiently, I grouped potential pools by running a hierarchical cluster analysis using a single linkage fusion strategy. I segregated clusters based on a threshold of 1/4 of a mile, ensuring that no two pools in a cluster will be more than 1/4 of a mile apart. This threshold was arbitrary, but seemed reasonable to keep pool clusters feeling contiguous.

```{r priority, include=FALSE, error=FALSE, warning=FALSE, message=FALSE}
# Create a priority index based on whether the cluster includes a pool that is 
# close to a certifed pool, the number of pools in the cluster, and the
# average distance to an impervious feature in the cluster in tens of meters.
# Join to filtered potential pool object.
cluster.prios <- cm.pp.f %>%
  dplyr::group_by(clust) %>%
  dplyr::summarise(
    n = n(),
    town = paste0(unique(TOWN)[1]),
    bio.flag = any(bio.flag == TRUE),
    # Aggregating to tens of meters smooths the ranking a bit. No reason to
    # privilege one pool over another based on a trivial distance difference.
    avg.dtf = round((mean(unlist(dtf))) / 10)) %>%
  dplyr::mutate(
    rank.b = dplyr::dense_rank(desc(bio.flag)),
    rank.n = dplyr::dense_rank(desc(n)),
    rank.dtf = dplyr::dense_rank(avg.dtf),
    prio = dplyr::dense_rank(
      dplyr::dense_rank(desc(bio.flag)) +
      dplyr::dense_rank(desc(n)) +
      dplyr::dense_rank(avg.dtf)) / 3)
# Add cluster centroid for reference
cluster.prios$centroid <- sf::st_centroid(cluster.prios$geometry)
cluster.prios <- sf::st_drop_geometry(cluster.prios)
cm.pp.f <- dplyr::left_join(
  cm.pp.f, cluster.prios[c("clust", "prio")], by = "clust")
```

Finally, I computed a prioritization index for each potential pool cluster based on an equal weighting between dense ranks over whether the cluster contained a pool within the dispersal area of a certified pool, the average distance to the nearest impervious feature among pools in the cluster, and the number of pools in the cluster. 

## Discussion

The most obvious caveat to the above analysis is the old adage "garbage in, garbage out." Because it combines data from different time periods, some several decades in the past, results may not reflect current conditions. Wetlands and vernal pools may have shifted boundaries, disappeared, or been developed over since the data were derived. There may also be bias in the potential pool data itself, since it was based on visual inspection of aerial photography, which is open to human error. The NHESP was able to validate that the false positive rate was low in trials (less than 3%), but it was unable to validate the false negative rate (Burne, 2001). It could be quite high because pools below ~50-60 feet in diameter could not be reliably identified (Burne, 2001).

The more critical caveat is that this analysis is very back-of-the-envelope. Proxy metrics, thresholds, and index design were informed by the literature where possible, but were fundamentally arbitrary and unvalidated.

## Results

`r nrow(cm.pp[cm.pp$dtw > 150, ])` potential pools (or `r scales::percent(round(nrow(cm.pp[cm.pp$dtw > 150, ]) / nrow(cm.pp), digits = 2))`) were excluded because they were not in close enough proximity to a wetland to fall under WPA jurisdiction.

Within the remaining `r nrow(cm.pp.f)` pools, I produced a map that citizen groups can use to identify hot spots of high value potential pools (coded in a heat map with more yellow colors indicating higher value). To validate that outcomes seemed reasonable, I also produced a zoomed in map and plotted potential pools, certified pools, potential pools removed from the data set, wetlands, and impervious features. This confirmed that removed pools were far from wetlands and high value pools tended to be clustered and close to certified pools.

```{r results_map, echo=FALSE, fig.height=2, fig.align='center', error=FALSE, warning=FALSE, message=FALSE}

# Set graphical parameters
par(mfrow = c(1, 2), mar = c(1, 1, 1, 1))

# Set extent for overall map.
ext1 <- c(xmin = 185280.1, ymin = 900004, xmax = 202250.2, ymax = 914339.6)

# Add town centroid for label plotting. Create plot of potential pools versus
# wetlands in all towns.
cm.towns <- cbind(cm.towns, st_coordinates(st_centroid(cm.towns)))
plot.all <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = sf::st_crop(cm.towns, ext1), fill = "white") +
  ggplot2::geom_sf(data = sf::st_crop(cm.wetlands, ext1), fill = "blue") +
  ggplot2::geom_sf(
    data = sf::st_crop(cm.pp.f, ext1), aes(color = prio), pch = 20, cex = 5) +
  ggplot2::scale_color_gradient(
    low="darkred", high="yellow", trans = "reverse", name = "Pool\nPrio") + 
  ggplot2::geom_label(
    data = sf::st_crop(cm.towns, ext1),
    aes(X, Y, label = TOWN),
    size = 4,
    fontface = "bold") +
  ggplot2::theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    rect = element_blank(),
    legend.margin = margin( 0, 0, 0, 0),
    legend.box.margin = margin(0, 0, 0, 0),
    plot.margin = unit(c(.5, 0, 0, 0), "cm"))

# Set extent for zoomed in map.
ext2 <- c(xmin = 196687.2, ymin = 903000, xmax = 202250.2, ymax = 906000)

# Create object combining deleted and certified pools.
cm.pp.del <- cm.pp[cm.pp$dtw > 150, ]
cm.pp.del$type <- "Removed"
cm.pp.del$pch <- 13
cm.pp.del$col <- "black"
cm.pp.del$cex <- 5
cm.pp.del <- cm.pp.del[, c("type", "pch", "col", "cex")]
cm.cp$type <- "Certified"
cm.cp$pch <- 20
cm.cp$col <- "blue"
cm.cp$cex <- 5
cm.cp <- cm.cp[, c("type", "pch", "col", "cex")]
cm.ep <- rbind(cm.pp.del, cm.cp)

# Create plot of zoomed in map area of Hudson
plot.zoom <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = sf::st_crop(cm.towns, ext2), fill = "white") +
  ggplot2::geom_sf(data = sf::st_crop(cm.wetlands, ext2), fill = "lightblue") +
  ggplot2::geom_sf(data = sf::st_crop(cm.tiles.f, ext2), fill = "lightgrey") +
  ggplot2::geom_sf(
    data = sf::st_crop(cm.pp.f, ext2), aes(color = prio), pch = 20, cex = 5) +
  ggplot2::scale_color_gradient(
    low="darkred", high="yellow", trans = "reverse", guide = "none") + 
  ggplot2::geom_sf(
    data = sf::st_crop(cm.ep, ext2),
    aes(shape = factor(type)), size = 2) +
  scale_shape_manual(values = c(17, 13)) +
  ggplot2::theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    rect = element_blank(),
    legend.margin = margin( 0, 0, 0, 0),
    legend.box.margin = margin(0, 0, 0, 0),
    legend.title = element_blank(),
    legend.position = "top",
    plot.margin = unit(c(.5, 0, 0, 0), "cm"))

# Plot maps side-by-side
cowplot::plot_grid(plot.all, plot.zoom, labels = c("Overview", "Zoomed In Region"))
```

## Conclusion

In the future, I'd like to expand this analysis in several ways. Covering the entire state would be valuable and would simply require more computational power than my personal laptop. I'd also like to move beyond the MassGIS data based on aerial photography, which, as mentioned, likely has high false negative rates. Lidar mapping has been shown to accurately identify vernal pools and could be used to expand my data set (Qiusheng, Lane, & Hongxing, 2014). 

But, most critically, I'd like to move beyond unvalidated back-of-the-envelope indexing and create a statistical model to predict which potential pools are likely to be strong certification candidates. To do this, I would require data on pools that were previously uncertified, submitted for review, and the certification outcome and WPA jurisdiction outcome of that review. Perhaps in the future such data could be compiled in partnership with citizen groups.

## Reference List

Burne, R.M. (2001) [Massachusetts Aerial Photo Survey of Potential Vernal Pools](https://www.mass.gov/doc/massachusetts-aerial-photo-survey-of-potential-vernal-pools/download) (2001). Natural Heritage & Endangered Species Program, Massachusetts Division of Fisheries and Wildlife.

Burne, R.B. & Griffin, C.R. (2005) Protecting vernal pools: a model from Massachusetts, USA. *Wetlands ecology and management*. Vol. 13 (3), 367-375. https://doi.org/10.1007/s11273-004-7528-3

Massachusetts Division Of Fisheries & Wildlife. (2009) [Guidelines for The Certification of Vernal Pool Habitat](https://www.mass.gov/doc/guidelines-for-the-certification-of-vernal-pool-habitat/download) [Accessed January 28, 2024].

McDonough, C. & Paton, P. (2010) Salamander Dispersal Across a Forested Landscape Fragmented by a Golf Course. *The Journal of Wildlife Management*. Vol. 71 (4), 1163-1169. https://doi.org/10.2193/2006-380

Qiusheng, W., Lane, C., & Hongxing, L. (2014) An Effective Method for Detecting Potential Woodland Vernal Pools Using High-Resolution LiDAR Data and Aerial Imagery. *Remote Sensing*. Vol. 6 (11), 11444-11467. https://doi.org/10.3390/rs61111444

\newpage
## Working Style Assessment
